Detect Pharmaceutical Quality Deviations with MS, Raman and Statistics

### 2025-02-13 dev/demo_ms_raman_fusion_bsa.Rmd
Major Update Class Names

The individual processing steps are applied according to *ProcessingSettings*. Below, the integration of the TIC is shown using the dedicated *ProcessingSettings* after loading the TIC chromatograms from the analysis files and smoothing using a moving average approach.

```{r load-chroms}
# Loads the TIC chromatograms
ms$run(MassSpecSettings_LoadChromatograms_StreamFind(chromatograms = "TIC"))
ms$analyses$has_chromatograms
```

```{r smooth-chroms}
# Smoothing of the TIC chromatogram for improving peak finding
ms$run(MassSpecSettings_SmoothChromatograms_movingaverage(windowSize = 5))
```

@@

# Settings for finding the local maxima in the loaded TIC chromatograms
ps_maxima <- MassSpecSettings_FindChromPeaks_LocalMaxima(
  minWidth = 3,
  maxWidth = 5,
  minHeight = 30E6
)

@@

According to the MS1 spectra, the *m/z* from 1200 to 2000 can be added to the dedicated *ProcessingSettings* for deconvolution, as shown below. The retention time dimensions for extraction of spectra is based on the local peak maxima shown above.

@@

# Extracts the spectra for the defined *m/z* range and local peak maxima
ms$run(
  MassSpecSettings_LoadSpectra_chrompeaks(
    mzmin = 1200,
    mzmax = 2000,
    levels = 1,
    minIntensity = 10
  )
)

# Calculates the charges for the extracted spectra
ms$run(
  MassSpecSettings_CalculateSpectraCharges_StreamFind(
    onlyTopScans = TRUE,
    topScans = 6,
    roundVal = 10,
    relLowCut = 0.2,
    absLowCut = 0,
    top = 10
  )
)

# Deconvolutes the charges for the extracted spectra
ms$run(MassSpecSettings_DeconvoluteSpectra_StreamFind())

ms$run(
  MassSpecSettings_SmoothSpectra_movingaverage(
    windowSize = 15
  )
)

# The binning is applied to ensure that all spectra have the same dimensions
# (i.e., mass-time and intensity pair)
ms$run(
   MassSpecSettings_BinSpectra_StreamFind(
    binNames = c("rt", "mass"),
    binValues = c(5, 20),
    byUnit = TRUE,
    refBinAnalysis = 1
  )
)

ms$run(
  MassSpecSettings_AverageSpectra_StreamFind(
    by = "replicates",
    weightedAveraged = FALSE
  )
)

ms$run(MassSpecSettings_NormalizeSpectra_minmax())

ms$run(MassSpecSettings_NormalizeSpectra_blockweight())
```

@@

```{r raman-workflow}
raman_workflow <- list(
  RamanSettings_BinScans_StreamFind(
    binNames = "rt",
    binValues = 20,
    byUnit = FALSE
  ),

  RamanSettings_SubtractSpectraSection_native(
    sectionVal = "rt",
    sectionWindow = c(0, 200)
  ),

  RamanSettings_DeleteSpectraSection_native(
    min = -100,
    shiftmax = 315
  ),

  RamanSettings_SmoothSpectra_savgol(
    fl = 11,
    forder = 2,
    dorder = 0
  ),

  RamanSettings_DeleteSpectraSection_StreamFind(
    shiftmin = 315,
    shiftmax = 330
  ),

  RamanSettings_DeleteSpectraSection_StreamFind(
    shiftmin = 2300,
    shiftmax = 2600
  ),

  RamanSettings_CorrectSpectraBaseline_baseline_als(
    lambda = 3,
    p = 0.015,
    maxit = 10
  ),

  RamanSettings_DeleteSpectraSection_StreamFind(
    rtmin = 0,
    rtmax = 420
  ),

  RamanSettings_DeleteSpectraSection_StreamFind(
    rtmin = 430,
    rtmax = 900
  ),

  RamanSettings_AverageSpectra_StreamFind(
    by = "replicates",
    weightedAveraged = FALSE
  ),

  RamanSettings_NormalizeSpectra_minmax(),

  RamanSettings_NormalizeSpectra_blockweight()
)

@@

```{r pca-run}
# Note that scale is set to FALSE to avoid scaling the data
# as blockweight normalization was applied for each dataset
stat$run(
  StatisticSettings_MakeModel_pca_mdatools(
    center = TRUE,
    ncomp = 2,
    alpha = 0.05,
    gamma = 0.02
  )
)
summary(stat$model) # summary method is from mdatools
```

@@

```{r mcr-run}
# Note that the model in the engine will be overwritten by the MCR model
stat$run(StatisticSettings_MakeModel_mcrpure_mdatools(ncomp = 2, offset = 0.05))
summary(stat$model)
```

>>>>

The individual processing steps are applied according to *ProcessingStep*. Below, the integration of the TIC is shown using the dedicated *ProcessingStep* after loading the TIC chromatograms from the analysis files and smoothing using a moving average approach.

```{r load-chroms}
# Loads the TIC chromatograms
ms$run(MassSpecMethod_LoadChromatograms_StreamFind(chromatograms = "TIC"))
ms$analyses$has_results_chromatograms
```

```{r smooth-chroms}
# Smoothing of the TIC chromatogram for improving peak finding
ms$run(MassSpecMethod_SmoothChromatograms_movingaverage(windowSize = 5))
```

@@

# Settings for finding the local maxima in the loaded TIC chromatograms
ps_maxima <- MassSpecMethod_FindChromPeaks_LocalMaxima(
  minWidth = 3,
  maxWidth = 5,
  minHeight = 30E6
)

@@

According to the MS1 spectra, the *m/z* from 1200 to 2000 can be added to the dedicated *ProcessingStep* for deconvolution, as shown below. The retention time dimensions for extraction of spectra is based on the local peak maxima shown above.

@@

# Extracts the spectra for the defined *m/z* range and local peak maxima
ms$run(
  MassSpecMethod_LoadSpectra_chrompeaks(
    mzmin = 1200,
    mzmax = 2000,
    levels = 1,
    minIntensity = 10
  )
)

# Calculates the charges for the extracted spectra
ms$run(
  MassSpecMethod_CalculateSpectraCharges_StreamFind(
    onlyTopScans = TRUE,
    topScans = 6,
    roundVal = 10,
    relLowCut = 0.2,
    absLowCut = 0,
    top = 10
  )
)

# Deconvolutes the charges for the extracted spectra
ms$run(MassSpecMethod_DeconvoluteSpectra_StreamFind())

ms$run(
  MassSpecMethod_SmoothSpectra_movingaverage(
    windowSize = 15
  )
)

# The binning is applied to ensure that all spectra have the same dimensions
# (i.e., mass-time and intensity pair)
ms$run(
   MassSpecMethod_BinSpectra_StreamFind(
    binNames = c("rt", "mass"),
    binValues = c(5, 20),
    byUnit = TRUE,
    refBinAnalysis = 1
  )
)

ms$run(
  MassSpecMethod_AverageSpectra_StreamFind(
    by = "replicates",
    weightedAveraged = FALSE
  )
)

ms$run(MassSpecMethod_NormalizeSpectra_minmax())

ms$run(MassSpecMethod_NormalizeSpectra_blockweight())
```

@@

```{r raman-workflow}
raman_workflow <- list(
  RamanMethod_BinScans_StreamFind(
    binNames = "rt",
    binValues = 20,
    byUnit = FALSE
  ),

  RamanMethod_SubtractSpectraSection_native(
    sectionVal = "rt",
    sectionWindow = c(0, 200)
  ),

  RamanMethod_DeleteSpectraSection_native(
    min = -100,
    shiftmax = 315
  ),

  RamanMethod_SmoothSpectra_savgol(
    fl = 11,
    forder = 2,
    dorder = 0
  ),

  RamanMethod_DeleteSpectraSection_StreamFind(
    shiftmin = 315,
    shiftmax = 330
  ),

  RamanMethod_DeleteSpectraSection_StreamFind(
    shiftmin = 2300,
    shiftmax = 2600
  ),

  RamanMethod_CorrectSpectraBaseline_baseline_als(
    lambda = 3,
    p = 0.015,
    maxit = 10
  ),

  RamanMethod_DeleteSpectraSection_StreamFind(
    rtmin = 0,
    rtmax = 420
  ),

  RamanMethod_DeleteSpectraSection_StreamFind(
    rtmin = 430,
    rtmax = 900
  ),

  RamanMethod_AverageSpectra_StreamFind(
    by = "replicates",
    weightedAveraged = FALSE
  ),

  RamanMethod_NormalizeSpectra_minmax(),

  RamanMethod_NormalizeSpectra_blockweight()
)

@@

```{r pca-run}
# Note that scale is set to FALSE to avoid scaling the data
# as blockweight normalization was applied for each dataset
stat$run(
  StatisticMethod_MakeModel_pca_mdatools(
    center = TRUE,
    ncomp = 2,
    alpha = 0.05,
    gamma = 0.02
  )
)
summary(stat$model) # summary method is from mdatools
```

@@

```{r mcr-run}
# Note that the model in the engine will be overwritten by the MCR model
stat$run(StatisticMethod_MakeModel_mcrpure_mdatools(ncomp = 2, offset = 0.05))
summary(stat$model)
```

### 2025-02-13 dev/demo_paracetamol_fusion.Rmd
Major Update Class Names

```{r ms-workflow}
ms_workflow <- list(
  MassSpecSettings_LoadSpectra_StreamFind(levels = 1, mzmax = 300, rtmax = 120),
  MassSpecSettings_BinSpectra_StreamFind(binNames = c("rt", "mz"), binValues = c(5, 2), refBinAnalysis = 1),
  MassSpecSettings_NormalizeSpectra_minmax(),
  MassSpecSettings_AverageSpectra_StreamFind(),
  MassSpecSettings_SubtractBlankSpectra_StreamFind(negativeToZero = TRUE),
  MassSpecSettings_NormalizeSpectra_blockweight()
)

@@

```{r raman-workflow}
raman_workflow <- list(
  RamanSettings_AverageSpectra_native(),
  RamanSettings_SubtractBlankSpectra_StreamFind(),
  RamanSettings_SmoothSpectra_savgol(fl = 21, forder = 4, dorder = 0),
  RamanSettings_DeleteSpectraSection_native(min = -100, max = 350),
  RamanSettings_DeleteSpectraSection_native(min = 1800, max = 2500),
  RamanSettings_CorrectSpectraBaseline_airpls(lambda = 45, differences = 1, itermax = 30),
  RamanSettings_NormalizeSpectra_minmax(),
  RamanSettings_NormalizeSpectra_blockweight()
)

@@

```{r pca-engine, fig.cap="PCA model (native plot from mdatools R package).", fig.height=9, warning=FALSE}
stat <- StatisticEngine$new(analyses = main_set)
# The alpha and gamma parameters are used to evaluate the model limits
stat$run(StatisticSettings_MakeModel_pca_mdatools(ncomp = 3, alpha = 0.01, gamma = 0.001))
plot(stat$model, showText = TRUE)
```

@@

```{r knn-engine}
stat2 <- StatisticEngine$new(analyses = train_set)
stat2$add_classes(labels)
stat2$run(StatisticSettings_MakeModel_knn(k = 3))
stat2$predict(test_set)
stat2$model$prediction$results
```

>>>>

```{r ms-workflow}
ms_workflow <- list(
  MassSpecMethod_LoadSpectra_StreamFind(levels = 1, mzmax = 300, rtmax = 120),
  MassSpecMethod_BinSpectra_StreamFind(binNames = c("rt", "mz"), binValues = c(5, 2), refBinAnalysis = 1),
  MassSpecMethod_NormalizeSpectra_minmax(),
  MassSpecMethod_AverageSpectra_StreamFind(),
  MassSpecMethod_SubtractBlankSpectra_StreamFind(negativeToZero = TRUE),
  MassSpecMethod_NormalizeSpectra_blockweight()
)

@@

```{r raman-workflow}
raman_workflow <- list(
  RamanMethod_AverageSpectra_native(),
  RamanMethod_SubtractBlankSpectra_StreamFind(),
  RamanMethod_SmoothSpectra_savgol(fl = 21, forder = 4, dorder = 0),
  RamanMethod_DeleteSpectraSection_native(min = -100, max = 350),
  RamanMethod_DeleteSpectraSection_native(min = 1800, max = 2500),
  RamanMethod_CorrectSpectraBaseline_airpls(lambda = 45, differences = 1, itermax = 30),
  RamanMethod_NormalizeSpectra_minmax(),
  RamanMethod_NormalizeSpectra_blockweight()
)

@@

```{r pca-engine, fig.cap="PCA model (native plot from mdatools R package).", fig.height=9, warning=FALSE}
stat <- StatisticEngine$new(analyses = main_set)
# The alpha and gamma parameters are used to evaluate the model limits
stat$run(StatisticMethod_MakeModel_pca_mdatools(ncomp = 3, alpha = 0.01, gamma = 0.001))
plot(stat$model, showText = TRUE)
```

@@

```{r knn-engine}
stat2 <- StatisticEngine$new(analyses = train_set)
stat2$add_classes(labels)
stat2$run(StatisticMethod_MakeModel_knn(k = 3))
stat2$predict(test_set)
stat2$model$prediction$results
```
